On the adhesivity of EGGS

### 2025-02-23 calco/main.tex
ritocchi minimi

@@

This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.

%\begin{notation}
Given a category $\X$ we will not distinguish notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$. We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of all arrows, monos and regular monos of $\X$, respectively.  Given an object $X$, we will denote by $?_X$ the unique arrow from an initial object into $X$ and by $!_X$ that  unique arrow from $X$ into a terminal one.
%\end{notation}

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive}.

In order to define it we need to introduce some terminology.  Let  $\X$ be a category. A subclass $\mathcal{A}$ of $\mor(\X)$ is called

@@

\begin{theorem}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category, then the following hold true:
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to A$ one preserving pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \hspace{5pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item if $\Y$ is a full subcategory of $\X$ closed in it under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ and  stable under pullbacks, pushout, containing all isomorphism and closed under composition and decomposition.
	\end{enumerate} 
\end{theorem}

@@

We can state some useful properties of $\mathcal{M}$-adhesive category.
\begin{restatable}{proposition}{regmono}\label[proposition]{prop:regmono}
	If $\X$ is $\mathcal{M}$-adhesive then the following are true:
	\begin{enumerate}
		\item every $\mathcal{M}$-pushout square is also a pullback;
		\item every arrow in $\mathcal{M}$ is a regular mono.
	\end{enumerate}
\end{restatable}
  [Proof in \Cref{regmono-proof}]

@@

In this section we recall the definition and some useful properties of \emph{kernel pairs} which will be useful in the next sections.

\noindent 
\parbox{11cm}{
\begin{definition}
    A \emph{kernel pair} for an arrow $f\colon A \to B$ is an object $K_f$ together with two arrows $\pi^1_f, \pi^2_f\colon K_f \rightrightarrows A$, denoted as $(K_f, \pi^1_f, \pi^2_f)$, such that the following square is a pullback.
\end{definition}}\hfill 
\parbox{2cm}{\xymatrix@R=15pt{K_f \ar[r]_{\pi^1_f} \ar[d]_{\pi^1_f}& A \ar[d]^f \\ A \ar[r]^{f}& B}}

\begin{remark}\label{prop:pairng_of_kernel_pairs_mono}
If $(K_f, \pi^1_f, \pi^2_f)$ is a a kernel pair for $f\colon X \to Y$ and a product of $X$ with itself exists, then the canonical arrow $\langle \pi^1_f, \pi^2_f\rangle \colon K_f \to X \times X$ is a mono.
\end{remark}

@@

Together with \Cref{lem:pb1}, the two previous remarks while elementary, allow us to prove the following result.

\begin{proposition}\label{cor:kermono}
	Let $f\colon X\to Y$ be an arrow and $m\colon Y\to Z$ be a mono. If
	$(K_f, \pi_f^1, \pi_f^2)$ is a kernel pair for $f\colon X\to Y$, then it is also a kernel pair for $m\circ f$.
\end{proposition}

@@

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be two arrows admitting kernel pairs and suppose that the solid part of the three squares below is given. Then there exists a a unique arrow $k_h: K_f \to K_g$ compliting them.
	\[\xymatrix{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z}\]

    Moreover, if the leftmost is a pullback, then also the other ones are so.
\end{restatable}
[Proof in \Cref{kpp-proof}]

@@

As a final step, we explore the relationship between regular epimorphisms and kernel pairs.

\begin{restatable}{proposition}{epic}\label{prop:reg_epi_coeq}
    Let $e\colon X \to Y$ be a regular epimorphism in a category $\X$ with a kernel pair $(K_e, \pi^1_e, \pi^2_e)$. Then, $e$ is the coequalizer of $\pi^1_e$ and $\pi^2_e$.
\end{restatable}
[Proof in \Cref{epic-proof}]

\begin{restatable}{corollary}{natepi}\label{cor:reg_epi_components_reg_epi_nat_trans}
    Let $\X$ be a category with pullbacks and $\phi\colon F \dot\to G$ be a natural transformation between two functors $F, G: \D \rightrightarrows \X$. If $\phi_d$ is a regular epi for every $d$ in $\D$, then $\phi$ is a regular epi.
\end{restatable}
[Proof in \Cref{natepi-proof}]

From the previous result we can now deduce that the class of regular epis is closed under colimits, as shown by the next lemma
\begin{restatable}{lemma}{epicol}\label{lemma:nat_trans_reg_epi_canonical_arrow_reg_epi}
    Let $F,G\colon \D\rightrightarrows \X$ be two diagrams, and suppose that $\X$ has all colimits of shape $\D$. Let $(X, \{x_d\}_{d \in \D})$ and $(Y, \{y_d\}_{d\in \S})$ be, respectively, the colimit of $F$ and $G$.  If $\phi\colon  F \to G$ is a natural transformation in which all components are regular epimorphisms, then, the arrow induced by $\phi$ from $X$ to $Y$ is a regular epimorphism too.
\end{restatable}
[Proof in \Cref{epicol-proof}]

@@

In this section we briefly recall the notion \emph{$\X$-hypergraph}. In order to do it is necessary to have a monad $(-)^\star$ playing a role analogous to the usual \emph{Kleene star}, also known as \emph{list monad}, $(-)^\star\colon \Set\to \Set$ sending a set to the free monoid on it \cite{sakarovitch2009elements,Wadler95}. We will recall some of its proprieties.

\begin{proposition}
	The following facts are true:
	\begin{enumerate}
		\item for every set $X$ and $n\in \mathbb{N}$ there are arrows $v_{n}\colon X^n\to X^\star $ such that $(X^\star, \{v_{n}\}_{n\in \mathbb{N}})$ is a coproduct;
		\item for every arrow $f\colon X\to Y$, $f^\star$ is the coproduct of the family $\{f^n\}_{n\in \mathbb{N}}$;
		\item $(-)^\star$ preserves all \emph{connected limits} \cite{carboni1995connected}, in particular it preserves all pullbacks and equalizers.
	\end{enumerate}
\end{proposition}

@@

\begin{remark}\label[remark]{rem:length}Notice that, $1^\star$ can be canonically identified with $\mathbb{N}$, thus for every set $X$, the arrow $!_\X\colon X\to 1$ induces a \emph{length function} $!^\star_X\colon X^\star \to \mathbb{N}$ which simply sends a word length.
\end{remark}

@@

We will start this section with the definition of hypergraphs and we will see how to label them with an algebraic signature.

\begin{definition}An \emph{hypergraph} is a 4-uple $\mathcal{G}:=(E_\mathcal{G}, V_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G})$ made by two sets $E_\mathcal{G}$ and $V_\mathcal{G}$, called respectively the set of \emph{hyperedges} and \emph{nodes}, plus a pair of \emph{source} and \emph{target} arrows  $s_\mathcal{G}, t_\mathcal{G}\colon E_\mathcal{G}\rightrightarrows V_\mathcal{G}^\star$. A \emph{hypergraph morphism} $(E_\mathcal{G}, V_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G})\to (E_\mathcal{H}, V_\mathcal{H}, s_\mathcal{H}, t_\mathcal{H})$ is a pair $(h,k)$ of functions $h\colon E_\mathcal{G}\to E_\mathcal{H}$, $k\colon V_\mathcal{G}\to V_\mathcal{H}$ such that the following diagrams commute.

	\[\xymatrix{ E_{\mathcal{G}} \ar[d]_{h} \ar[r]^{s_{\mathcal{G}}}& V^\star_{\mathcal{G}}  \ar[d]^{k^\star}& E_{\mathcal{G}} \ar[r]^{t_{\mathcal{G}}} \ar[d]_{h} & V^\star_{\mathcal{G}} \ar[d]^{k^\star}  \\ E_{\mathcal{G}} \ar[r]_{s_{\mathcal{H}}} & V^\star_{\mathcal{H}} & E_{\mathcal{G}} \ar[r]_{t_{\mathcal{H}}} & V^\star_{\mathcal{H}} }\]
	We define $\hyp$ to be the resulting category.
\end{definition}

Let $\pro^\star$ be the functor sending $X$ to $X^\star\times X^\star$, then we can present $\hyp$ as a comma category.
\begin{proposition}\label[proposition]{prop:com}
	$\hyp$ is isomorphic to $\comma{\id{\Set}}{\pro^\star}$
\end{proposition}

@@

\begin{example}Since the initial object of $\catname{Set}$ is the empty set,  $\Delta_{\Set}(X)$ is the hypergraph which has $X$ as set of nodes and $\emptyset$ as set of hyperedges and $?_X$ as both source and target function.
\end{example}

@@

\begin{definition}Let $\catname{H}$ be the category in which:
	\begin{itemize}
		\item the set of objects is given by $ (\mathbb{N}\times \mathbb{N}) \cup \{\bullet\}$
		\item arrows are given by the identities $\id{k,l}$ and $\id{\bullet}$ and exactly $k+l$ arrows $f_i\colon (k,l)\rightarrow \bullet$, where $i$ ranges from $0$ to $k+l-1$;
		\item composition is defined simply putting, for every $f_i\colon (k,l)\rightarrow \bullet$:
		\begin{equation*}
			f_i=f_i\circ \id{k,l} \qquad f_i = \id{\bullet}\circ f_i 
		\end{equation*}
	\end{itemize}
\end{definition}

The idea now is that for every functor $F\colon \catname{H}\to \catname{Set}$ we can define
\[E_F:=\sum_{k,l\in \mathbb{N}}F(k,l)\]

Now, for every $k$, $l$, $i$ and $j$ in $\mathbb{N}$ with $i< k$ and $j< l$ we can define $s^F_{k,l}\colon F(k,l)\to F(\bullet)^k$,  $t^F_{k,l}\colon F(k,l)\to F(\bullet)^l$ as the unique ones fitting in the diagrams below, where the vertical arrows are the projections.
\[\xymatrix{F(k,l)  \ar@{.>}[r]^{s^F_{k,l}} \ar[dr]_{F(f_i)}& F(\bullet)^{k} \ar[d]^{\pi^F_{k,i}} & F(k,l) \ar@{.>}[r]^{t^F_{k,l}} \ar[dr]_{F(f_{k+j})} & F(\bullet)^{l} \ar[d]^{\pi^F_{l,j}} \\ & F(\bullet) && F(\bullet)}\]

In turn, these arrows allow us to consider
$s_F, t_F\colon E_F\rightrightarrows F(\bullet)^{\star}$ as the unique arrows fitting in the diagrams below, where the vertical arrows are coprojections.
\[\xymatrix{F(k,l) \ar[d]_{a^F_{k,l}}  \ar[r]^{s^{F}_{k,l}}& F(\bullet)^{k} \ar[d]^{b^F_{k}} & F(k,l) \ar[d]_{a^F_{k,l}}  \ar[r]^{t^{F}_{k,l}}& F(\bullet)^{l} \ar[d]^{b^F_{l}}\\ E_F \ar@{.>}[r]_-{s_F}& F(\bullet)^\star & E_F \ar@{.>}[r]_-{t_F}& F(\bullet)^\star}\]

@@

Our interest intohypergraph stems fromm their use as a graphical representation of algebraic terms. We thus need a way to label hyperedges with symbols taken from a given algebraic signature.

@@

\begin{example}\label[example]{exa_3} Let $\Sigma=(O_\Sigma, \ari_\Sigma)$ be an algebraic signature in $\Set$. This simply amount to a set of \emph{operations} with an associated natural number, called \emph{arity}. 	For instance let $\Sigma_G$ be the signature of groups, then $\mathcal{G}^{\Sigma_G}$ can be depicted as:
	\begin{center}
		\begin{tikzpicture}
			\node[circle,fill=black,inner sep=0pt,minimum size=6pt,label=above:{$v$}] (V) at (0,0) {};
			\node(E)at(-2, 0.4){$e$};
			\node(M)at(0, 2.15){$\cdot$};
			\node(I)at(2, 0.5){$(-)^{-1}$};
			\draw[->-=.5](-1.75,0)--(V)node[pos=0.5, above,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](V)..controls(-0.5,0.5)and(-0.8,1)..(-0.25,1.6)node[pos=0.5, right,font=\fontsize{7}{0}\selectfont]{$2$};
			\draw[->-=.5](V)..controls(-1,0.6)and(-1,1.1)..(-0.25,1.9)node[pos=0.5, left,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](0.25,1.75)..controls(0.8,0.8)and(0.5,0.5)..(V)node[pos=0.5, right,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](V)--(1.75,0)node[pos=0.5, above,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](2.25,0)..controls(3.5,0)and(2.5,-2)..(V)node[pos=0.5, below,font=\fontsize{7}{0}\selectfont]{$2$};
			\draw[rounded corners] (-2.25, -0.25) rectangle (-1.75, 0.25) {};
			\draw[rounded corners] (-0.25, 1.5) rectangle (0.25, 2) {};
			\draw[rounded corners] (2.25, -0.25) rectangle (1.75, 0.25) {};
		\end{tikzpicture}
	\end{center}
\end{example}

\Cref{cor:mono} and \Cref{thm:slice-functors} give us immediately an adhesivity result for $\hyp_{\Sigma}$ and a characterization of monomorphisms in it.

@@

\begin{remark}\label[remark]{rem:mono2}By \Cref{rem:label}, we know that if $\mathcal{G}$ is a term graph then $t_{\mathcal{G}}=v_1\circ \tau_{\mathcal{G}}$, where $v_1$ is the coprojection of $V_{\mathcal{G}}$ into $V^\star_{\mathcal{G}}$.  Notice, moreover that since $t_{\mathcal{G}}$ is mono then $\tau_{\mathcal{G}}$ is mono too.
\end{remark}

@@

\begin{proposition}\label[proposition]{term:left}The forgetful functor $U_{\tg}$ has a left adjoint $\Delta_{\tg}$.
\end{proposition}
\begin{proof}
	This follows noticing that, for every object $X$, $\Delta_{\Sigma}(X)$ is a term graph.   
\end{proof}

We canlist some categorical properties of $\tg$

@@

Our result opens two threads of research. The first is to use the quasi-adhesivity of EGGS in order to
model their rewriting via the double-pushout (DPO) approach. This seems to be straightforward, since 
the rules adopted in the literature of EGGS appears to be span of regular monos, and such rules
perfectly fit the mold of rewriting on quasi-adhesive categories. 
For example, the equivalence $x \div x = 1$, taken from the introductory example in~\cite{xxx}, 
can be modelled as the rule
\[to draw DPO rule\]

It still needs to be investigated what concurrency and termination, the key properties for DPO rewriting
on adhesive categories, 
mean in the context of EGGS. More interestingly, another venue for development is using the adhesive 
machinery to extend the EGGS formalism. In fact, most of the results
presented here for hyper-graphs can be generalised to hierarchical hyper-graphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges that is useful for adding structural information, such as encapsulation and sandboxing~\cite{xxx}.

>>>>

\newcommand{\commentato}[1]{ {} }

@@

This  section briefly recalls \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}. 
%
%\begin{notation}
Given a category $\X$ we do not distinguish notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$. We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of all arrows, monos and regular monos of $\X$, respectively.  Given an object $X$, we  denote by $?_X$ the unique arrow from an initial object into $X$ and by $!_X$ that  unique arrow from $X$ into a terminal one.
%\end{notation}

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive},
%
and to define it we introduce some terminology.  Let  $\X$ be a category. A subclass $\mathcal{A}$ of $\mor(\X)$ is

@@

\begin{theorem}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category, then the following hold true:
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to A$ one preserving pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \hspace{5pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item if $\Y$ is a full subcategory of $\X$ closed in it under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ and  stable under pullbacks and pushout, containing all isomorphisms and closed under composition and decomposition.
	\end{enumerate} 
\end{theorem}

@@

We can state some useful properties of $\mathcal{M}$-adhesive category.

\begin{restatable}{proposition}{regmono}\label{prop:regmono}
	If $\X$ is $\mathcal{M}$-adhesive then the following are true:
	\begin{enumerate}
		\item every $\mathcal{M}$-pushout square is also a pullback;
		\item every arrow in $\mathcal{M}$ is a regular mono.
	\end{enumerate}
\end{restatable}
  [Proof in \Cref{regmono-proof}]

@@

In this section we recall the definition and some properties of \emph{kernel pairs}.
%which will be useful in the next sections.

\noindent 
\parbox{11cm}{
\begin{definition}
    A \emph{kernel pair} for an arrow $f\colon A \to B$ is an object $K_f$ together with two arrows $\pi^1_f, \pi^2_f\colon K_f \rightrightarrows A$, denoted as $(K_f, \pi^1_f, \pi^2_f)$, such that the square aside is a pullback.
\end{definition}}\hfill 
\parbox{2cm}{\xymatrix@R=15pt{K_f \ar[r]_{\pi^2_f} \ar[d]_{\pi^1_f}& A \ar[d]^f \\ A \ar[r]^{f}& B}}

\smallskip
\begin{remark}\label{prop:pairng_of_kernel_pairs_mono}
If $(K_f, \pi^1_f, \pi^2_f)$ is a a kernel pair for $f\colon X \to Y$ and a product of $X$ with itself exists, then the canonical arrow $\langle \pi^1_f, \pi^2_f\rangle \colon K_f \to X \times X$ is a mono.
\end{remark}

@@

Together with \Cref{lem:pb1}, the previous remarks allow us to prove the following result.

\begin{proposition}\label{cor:kermono}
	Let $f\colon X\to Y$ be an arrow and $m\colon Y\to Z$ a mono. If
	$(K_f, \pi_f^1, \pi_f^2)$ is a kernel pair for $f\colon X\to Y$, then it is also a kernel pair for $m\circ f$.
\end{proposition}

@@

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be two arrows admitting kernel pairs and suppose that the solid part of the three squares below is given. Then there exists a a unique arrow $k_h: K_f \to K_g$ completing them.
	\[\xymatrix{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z}\]

    Moreover, if the leftmost is a pullback, then also the other two are so.
\end{restatable}
[Proof in \Cref{kpp-proof}]

@@

As a final step, we explore the link between regular epis and kernel pairs.

\begin{restatable}{proposition}{epic}\label{prop:reg_epi_coeq}
    Let $e\colon X \to Y$ be a regular epi in a category $\X$ with a kernel pair $(K_e, \pi^1_e, \pi^2_e)$. Then, $e$ is the coequalizer of $\pi^1_e$ and $\pi^2_e$.
\end{restatable}
[Proof in \Cref{epic-proof}]

\begin{restatable}{corollary}{natepi}\label{cor:reg_epi_components_reg_epi_nat_trans}
    Let $\X$ be a category with pullbacks and $\phi\colon F \dot\to G$ a natural transformation between functors $F, G: \D \rightrightarrows \X$. If $\phi_d$ is a regular epi for every $d$ in $\D$, then $\phi$ is a regular epi.
\end{restatable}
[Proof in \Cref{natepi-proof}]

From the previous result we deduce that the class of regular epis is closed under colimits.
% as shown by the next lemma

\begin{restatable}{lemma}{epicol}\label{lemma:nat_trans_reg_epi_canonical_arrow_reg_epi}
    Let $F,G\colon \D\rightrightarrows \X$ be two diagrams, and suppose that $\X$ has all colimits of shape $\D$. Let $(X, \{x_d\}_{d \in \D})$ and $(Y, \{y_d\}_{d\in \S})$ be the colimits of $F$ and $G$, respectively.  If $\phi\colon  F \to G$ is a natural transformation whose components are regular epi, then the arrow induced by $\phi$ from $X$ to $Y$ is a regular epi too.
\end{restatable}
[Proof in \Cref{epicol-proof}]

@@

In this section we briefly recall the notion \emph{$\X$-hypergraph}. It is necessary to have a monad $(-)^\star$ playing a role analogous to the usual \emph{Kleene star}
 $(-)^\star\colon \Set\to \Set$, also known as \emph{list monad}, sending a set to the free monoid on it \cite{sakarovitch2009elements,Wadler95}. We recall some of its proprieties.

\begin{proposition}
	The following facts are true:
	\begin{enumerate}
		\item for every set $X$ and $n\in \mathbb{N}$ there are arrows $v_{n}\colon X^n\to X^\star $ such that $(X^\star, \{v_{n}\}_{n\in \mathbb{N}})$ is a coproduct;
		\item for every arrow $f\colon X\to Y$, $f^\star$ is the coproduct of the family $\{f^n\}_{n\in \mathbb{N}}$;
		\item $(-)^\star$ preserves all \emph{connected limits} \cite{carboni1995connected}, in particular it preserves pullbacks and equalizers.
	\end{enumerate}
\end{proposition}

@@

\begin{remark}\label[remark]{rem:length}Notice that $1^\star$ can be canonically identified with $\mathbb{N}$, thus for every set $X$ the arrow $!_\X\colon X\to 1$ induces a \emph{length function} $!^\star_X\colon X^\star \to \mathbb{N}$, which sends a word to its length.
\end{remark}

@@

We open this section with the definition of hypergraphs and we will see how to label them with an algebraic signature.

\begin{definition}An \emph{hypergraph} is a 4-uple $\mathcal{G}:=(E_\mathcal{G}, V_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G})$ made by two sets $E_\mathcal{G}$ and $V_\mathcal{G}$, called respectively the set of \emph{hyperedges} and \emph{nodes}, plus a pair of \emph{source} and \emph{target} arrows  $s_\mathcal{G}, t_\mathcal{G}\colon E_\mathcal{G}\rightrightarrows V_\mathcal{G}^\star$. A \emph{hypergraph morphism} $(E_\mathcal{G}, V_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G})\to (E_\mathcal{H}, V_\mathcal{H}, s_\mathcal{H}, t_\mathcal{H})$ is a pair $(h,k)$ of functions $h\colon E_\mathcal{G}\to E_\mathcal{H}$, $k\colon V_\mathcal{G}\to V_\mathcal{H}$ such that the following diagrams commute
	\[\xymatrix{ E_{\mathcal{G}} \ar[d]_{h} \ar[r]^{s_{\mathcal{G}}}& V^\star_{\mathcal{G}}  \ar[d]^{k^\star}& E_{\mathcal{G}} \ar[r]^{t_{\mathcal{G}}} \ar[d]_{h} & V^\star_{\mathcal{G}} \ar[d]^{k^\star}  \\ E_{\mathcal{G}} \ar[r]_{s_{\mathcal{H}}} & V^\star_{\mathcal{H}} & E_{\mathcal{G}} \ar[r]_{t_{\mathcal{H}}} & V^\star_{\mathcal{H}} }\]
	We define $\hyp$ to be the resulting category.
\end{definition}

Let $\pro^\star$ be the functor sending $X$ to $X^\star\times X^\star$: we can get $\hyp$ as a comma category.

\begin{proposition}\label[proposition]{prop:com}
	$\hyp$ is isomorphic to $\comma{\id{\Set}}{\pro^\star}$
\end{proposition}

@@

\begin{example}Since the initial object of $\catname{Set}$ is the empty set,  $\Delta_{\Set}(X)$ is the hypergraph which has $X$ as set of nodes, $\emptyset$ as set of hyperedges, and $?_X$ as source and target function.
\end{example}

@@

\begin{definition}Let $\catname{H}$ be the category isuch that
	\begin{itemize}
		\item the set of objects is $ (\mathbb{N}\times \mathbb{N}) \cup \{\bullet\}$;
		\item arrows are given by the identities $\id{k,l}$ and $\id{\bullet}$ and exactly $k+l$ arrows $f_i\colon (k,l)\rightarrow \bullet$, where $i$ ranges from $0$ to $k+l-1$;
		\item composition is defined by putting
		%\begin{equation*}
			$f_i=f_i\circ \id{k,l}$ and $f_i = \id{\bullet}\circ f_i$
		%\end{equation*}
		for every $f_i\colon (k,l)\rightarrow \bullet$.
	\end{itemize}
\end{definition}

The idea is that for every functor $F\colon \catname{H}\to \catname{Set}$ we can define
\[E_F:=\sum_{k,l\in \mathbb{N}}F(k,l)\]
%
Now, for every $k$, $l$, $i$ and $j$ in $\mathbb{N}$ with $i< k$ and $j< l$ we define $s^F_{k,l}\colon F(k,l)\to F(\bullet)^k$ and  $t^F_{k,l}\colon F(k,l)\to F(\bullet)^l$ as the unique arrows fitting in the diagrams below, where the vertical arrows are the projections
\[\xymatrix{F(k,l)  \ar@{.>}[r]^{s^F_{k,l}} \ar[dr]_{F(f_i)}& F(\bullet)^{k} \ar[d]^{\pi^F_{k,i}} & F(k,l) \ar@{.>}[r]^{t^F_{k,l}} \ar[dr]_{F(f_{k+j})} & F(\bullet)^{l} \ar[d]^{\pi^F_{l,j}} \\ & F(\bullet) && F(\bullet)}\]
%
In turn, these arrows allow us to consider
$s_F, t_F\colon E_F\rightrightarrows F(\bullet)^{\star}$ as the unique arrows fitting in the diagrams below, where the vertical arrows are coprojections
\[\xymatrix{F(k,l) \ar[d]_{a^F_{k,l}}  \ar[r]^{s^{F}_{k,l}}& F(\bullet)^{k} \ar[d]^{b^F_{k}} & F(k,l) \ar[d]_{a^F_{k,l}}  \ar[r]^{t^{F}_{k,l}}& F(\bullet)^{l} \ar[d]^{b^F_{l}}\\ E_F \ar@{.>}[r]_-{s_F}& F(\bullet)^\star & E_F \ar@{.>}[r]_-{t_F}& F(\bullet)^\star}\]

@@

Our interest for hypergraphs stems from their use as a graphical representation of algebraic terms. We thus need a way to label hyperedges with symbols taken from a signature.

@@

\begin{example}\label[example]{exa_3} Let $\Sigma=(O_\Sigma, \ari_\Sigma)$ be an algebraic signature in $\Set$. This simply amount to a set of \emph{operations} with an associated natural number, called \emph{arity}. 	For instance let $\Sigma_G$ be the signature of groups, then $\mathcal{G}^{\Sigma_G}$ can be depicted as
	\begin{center}
		\begin{tikzpicture}
			\node[circle,fill=black,inner sep=0pt,minimum size=6pt,label=above:{$v$}] (V) at (0,0) {};
			\node(E)at(-2, 0.4){$e$};
			\node(M)at(0, 2.15){$\cdot$};
			\node(I)at(2, 0.5){$(-)^{-1}$};
			\draw[->-=.5](-1.75,0)--(V)node[pos=0.5, above,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](V)..controls(-0.5,0.5)and(-0.8,1)..(-0.25,1.6)node[pos=0.5, right,font=\fontsize{7}{0}\selectfont]{$2$};
			\draw[->-=.5](V)..controls(-1,0.6)and(-1,1.1)..(-0.25,1.9)node[pos=0.5, left,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](0.25,1.75)..controls(0.8,0.8)and(0.5,0.5)..(V)node[pos=0.5, right,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](V)--(1.75,0)node[pos=0.5, above,font=\fontsize{7}{0}\selectfont]{$1$};
			\draw[->-=.5](2.25,0)..controls(3.5,0)and(2.5,-2)..(V)node[pos=0.5, below,font=\fontsize{7}{0}\selectfont]{$2$};
			\draw[rounded corners] (-2.25, -0.25) rectangle (-1.75, 0.25) {};
			\draw[rounded corners] (-0.25, 1.5) rectangle (0.25, 2) {};
			\draw[rounded corners] (2.25, -0.25) rectangle (1.75, 0.25) {};
		\end{tikzpicture}
	\end{center}
\end{example}

\Cref{cor:mono} and \Cref{thm:slice-functors} give us immediately an adhesivity result for $\hyp_{\Sigma}$ and a characterisation of monos in it.

@@

\begin{remark}\label[remark]{rem:mono2}By \Cref{rem:label}, we know that if $\mathcal{G}$ is a term graph then $t_{\mathcal{G}}=v_1\circ \tau_{\mathcal{G}}$, where $v_1$ is the coprojection of $V_{\mathcal{G}}$ into $V^\star_{\mathcal{G}}$.  Notice that since $t_{\mathcal{G}}$ is mono then $\tau_{\mathcal{G}}$ is mono too.
\end{remark}

@@

\begin{proposition}\label[proposition]{term:left}The forgetful functor $U_{\tg}$ has a left adjoint $\Delta_{\tg}$.
\end{proposition}
\begin{proof}
	This follows noticing that $\Delta_{\Sigma}(X)$ is a term graph for every object $X$.
\end{proof}

We can list some categorical properties of $\tg$

@@

Our result opens two threads of research. The first is to use the quasi-adhesivity of EGGs to
model their rewriting via the double-pushout (DPO) approach. This seems now easy, since 
the rules adopted in the literature of EGGs appears to be span of regular monos, and such rules
perfectly fit the mold of rewriting on quasi-adhesive categories. 
For example, the equivalence $x \div x = 1$, from the introductory example in~\cite{xxx}, 
can be modelled as the rule
\[to draw DPO rule\]

It still needs to be investigated what concurrency and termination, the key properties for DPO rewriting
on adhesive categories, 
mean in the context of EGGs. More interestingly, another venue for development is using the adhesive 
machinery to extend the EGGs formalism. In fact, most of the results
presented here for hyper-graphs can be generalised to hierarchical hyper-graphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges that is useful for adding structural information, such as encapsulation and sandboxing~\cite{xxx}.

Finally, we need to draw a comparison with an alternative categorical presentation for EGGs 
advanced in~\cite{ghica}. The proposal is quite different from our own. Simplifying, the key is to equip categories 
of trees with a lattice on hom-sets. Instead of going through 
such an enrichment, we chose to obtain closure via suitable requirements on morphisms, even if it 
seems that the former generalises the latter, at the expenses of a more complex machinery.
